SciPy 1.0---Fundamental Algorithms for Scientific Computing in Python

### 2019-01-14 scipy-1.0/paper.tex
Add SI Figures for Sparse Performance
* add / cite SI Figures for our sparse
performance improvements

* currently, it appears that the CSR to COO
performance improvement is not substantiated
by the asv results

SciPy's arrival at this point is surprising, and somewhat anomalous.
When started, the library had little funding, and was written mainly
by graduate students---many of them without a computer science education, and often without the
blessing of their advisors.  To even imagine that a small group of
``rogue'' student programmers could upend the already well-established
ecosystem of research software---backed by millions in funding, and
thousands of highly qualified engineers---was preposterous.

@@

In 1995, Jim Hugunin, a graduate student from MIT, wrote in the first
message on the Python Matrix Special Interest Group (Matrix-SIG)
mailing list\footnote{https://mail.python.org/pipermail/matrix-sig/1995-August/000002.html}:
\begin{quote}
There seems to be a fair amount of interest in the Python community
concerning the addition of numeric operations to Python.  My own desire is
to have as large a library of matrix based functions available as possible
(linear algebra, eigenfunctions, signal processing, statistics, etc.).  In
order to ensure that all of these libraries interoperate, there needs to
be agreement on a basic matrix object that can be used to represent arrays
of numbers.
\end{quote}
Over the next several months, conversations on that mailing
list by, among others, Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
Hinsen, and Guido van Rossum, led to the creation of an array object
that supported a high number of dimensions.  Jim Hugunin explained the utility
of Python for numerical computation\footnote{http://hugunin.net/papers/hugunin95numpy.html}:
\begin{quote}
I've used almost all of the available numerical languages at one time
or another over the past 8 years. One thing I've noticed is that over
time, the designers of these languages are steadily adding more of the
features that one would expect to find in a general-purpose
programming language.
\end{quote}
This is still a distinguishing feature of Python for science, and one of the
reasons why it has been so successful in the realm of data science: instead of
adding general features to a language designed for numerical and scientific
computing, here scientific features are added to a general purpose language
instead.  This broadens the scope of problems that can be addressed easily, the
sources of data that are readily accessible, and increases the size of the
community that develops general purpose tooling.

@@

Initial work on numerical computing in Python was driven by graduate
students, but soon larger research labs became increasingly engaged.
For example, Paul Dubois at Lawrence Livermore National Laboratory (LLNL) took over the
maintenance of Numeric and funded the writing of its
manual\footnote{http://www.numpy.org/\_downloads/numeric-manual.pdf}, and
%in the Program for Climate Model Diagnosis and Intercomparison at
%Lawrence Livermore National Laboratory built their Climate Data
%Analysis Tools (``an open-source set of Python modules and tools for
%accessing, manipulating, and plotting scientific data in general and
%climate data sets in
%particular''\footnote{https://mail.python.org/pipermail/numpy-discussion/2001-May/000645.html})
%on Numeric, the newly established Python array object.  Paul later
%became the maintainer for Numeric, and helped fund the writing of the
%first manual\footnote{https://conference.scipy.org/scipy2010/slides/travis_oliphant_keynote.pdf}
in 1998 the Space Telescope Science Institute (STScI), in charge of science
operations of the Hubble Space Telescope, decided to replace their
custom scripting language with Python\footnote{https://conference.scipy.org/scipy2011/slides/greenfield\_keynote\_astronomy.pdf}.

@@

% PhD Mayo Clinic, started faculty position at BYU in 2001
By the late 1990s, discussions appeared on Matrix-SIG,
expressing a desire for a complete scientific data analysis environment\footnote{https://conference.scipy.org/scipy2010/slides/travis\_oliphant\_keynote.pdf}.
Travis Oliphant, a PhD student at the Mayo Clinic,
released a number of packages\footnote{https://web.archive.org/web/19990125091242/http://oliphant.netpedia.net:80/}\footnote{https://web.archive.org/web/20001206213500/http://oliphant.netpedia.net:80/} that built on top of the Numeric array
package, and provided algorithms for signal processing, special
functions, sparse matrices, quadrature, optimization, Fast Fourier
Transforms, and more.  One of these packages, Multipack, was a set of
extension modules that wrapped Fortran and C libraries such as
ODEPACK, QUADPACK, and MINPACK, to solve and minimize nonlinear
equations, integrate differential equations, and fit splines (the
latter implemented by Pearu Peterson).  Robert Kern, then an
undergraduate student (and currently a SciPy core developer), provided
compilation instructions under Windows.
% 2001 PhD Tallinn Technical University
Around the same time, Pearu Peterson, a PhD student from Estonia,
released f2py\footnote{https://mail.python.org/pipermail/numpy-discussion/2000-September/000243.html}, a command line tool for binding Python and Fortran
codes, and wrote modules for linear algebra and interpolation.
% Graduated 1999, Duke
Eric Jones, while a graduate student at Duke, wrote a number of
packages to support his dissertation, including a parallel job
scheduler and genetic optimizer.
% Graduated 1998, PhD from Brown, 97–99 postdoc at Harvard Medical
% School; at Harvard ever since
%
% https://mail.python.org/pipermail/scipy-dev/2001-November/000088.html
%
Gary Strangman, a postdoctoral fellow at Harvard Medical School,
published several descriptive and inferential statistical routines\footnote{https://web.archive.org/web/20001022231108/http://www.nmr.mgh.harvard.edu/Neural\_Systems\_Group/gary/python.html}.

@@

With a rich programming environment and a numerical array object in
place, the time was ripe for the development of a full scientific
software stack.
In 2001, Eric Jones and Travis Vaught founded Enthought Scientific
Computing Solutions (now Enthought, Inc.) in Austin, Texas.  In an
effort to simplify the tool stack, they created the SciPy project,
centered around the SciPy library which would subsume all the
above-mentioned packages.
The new project quickly gained momentum, with a website and code
repository appearing in
February\footnote{https://web.archive.org/web/20010309040805/http://scipy.org:80/}, and
a mailing list announced in
June \footnote{https://mail.python.org/pipermail/scipy-dev/2001-June/000000.html}. By
August, a first release was announced\footnote{https://mail.python.org/pipermail/python-list/2001-August/106419.html}, an excerpt of which is shown in
Fig.~\ref{fig:announce-0.1}.
In September, the first documentation was
published\footnote{https://web.archive.org/web/20021013204556/http://www.scipy.org:80/scipy/site\_content/site\_news/docs\_released1}.
The first SciPy
workshop\footnote{https://mail.python.org/pipermail/numpy-discussion/2002-June/001511.html}
was held in September 2002 at CalTech---a single track, two day event with 50
participants, many of them developers of SciPy and surrounding libraries.

@@

As SciPy, the algorithms library of the ecosystem, began attracting the attention
of large research organizations,
the next generation of graduate students and postdocs were already exploring
other aspects of the computational environment.
In 2000, Prabhu Ramachandran, a PhD student at the Indian Institute of
Technology, started work on a 3D visualization application, based on
Wingware's C++ Visualization Toolkit, called
Mayavi\footnote{https://web.archive.org/web/20030731122452/http://pythonology.org/success\&story=mayavi}.
In 2001, Fernando Pérez, a graduate student at the University of
Colorado, Boulder, created the IPython interactive shell (that would
eventually blossom into Project Jupyter\cite{Kluyver:2016aa}), built into
IPP, an earlier attempt by Janko Hauser, and including code from
Nathan Gray's LazyPython.
% IPython hosted on SciPy:
% https://mail.python.org/pipermail/scipy-user/2003-March/001455.html
John Hunter, a postdoc
%John got his PhD in 2001, then a post-doc with John Milton (dept of
%Neurology, UChicago?) a couple of years; then research professor
%under Kurt Hecox, whereafter at Tradelink from 2006
at the University
of Chicago, liked the plotting functionality
available in MATLAB, but had problems accessing their laboratory's
license, which was governed by a hardware dongle.  In response, he
wrote a plotting library from scratch, and Matplotlib 0.1 was released
April 2003\footnote{https://mail.python.org/pipermail/python-list/2003-April/193167.html}.

@@

%In April 2006, Paul
%Dubois\footnote{https://mail.python.org/pipermail/numpy-discussion/2006-April/007487.html} wrote
%to the NumPy mailing list:
%\begin{quote}
%IEEE's magazine, Computing in Science and Engineering (CiSE), has asked me
%to put together a theme issue on the use of Python in Science and
%Engineering. I will write an overview to be accompanied by 3-5 articles [. . .]
%that show a diverse set of applications or
%tools, to give our readers a sense of whether or not Pyth%on might be useful
%in their own work.
%\end{quote}
In a May/June 2007 special issue of IEEE Computing in Science and
Engineering, Paul Dubois wrote\cite{dubois2007guest}:
\begin{quote}
LLNL now has many Python-based efforts built from scratch or wrapped around
legacy codes [. . .] hundreds of thousands
of lines of C++, Python, and Fortran 95, all working together just as we hoped,
doing compute-intensive calculations on massively parallel computers.
\end{quote}
While it is now more commonly accepted, Paul took the time to explain to
the 2007 reader that ``interpreted doesn't mean slow or only interactive.''
He also shares his own interest in Python for ``computational steering''
where ``Python serves as the input language to a scientific application, and the actual
computations are performed both in Python itself and in compiled extensions.''
%https://www.computer.org/csdl/mags/cs/2007/03/c3007.html
Using Python for computational steering was one of the earliest uses of Python
in large scientific processing pipelines.
In addition to Paul's overview there were also articles introducing
Python and SciPy \cite{oliphant2007python}, IPython \cite{perez2007ipython},
and  Matplotlib \cite{hunter2007matplotlib}.
In addition to these longer introductory articles, there was a diverse set of
research applications including
systems biology\cite{myers2007python},
astronomy\cite{greenfield2007reaching},
robotics\cite{krauss2007python},
nanophotonics\cite{bienstman2007python},
partial differential equations\cite{mardal2007using},
neuroimaging\cite{millman2007analysis},
geographic information systems\cite{shi2007python}, and
education\cite{backer2007computational, myers2007pythona}.

@@

By the middle to late 2000s, SciPy was starting to mature after a long phase of significant
growth and adoption.
The informal workshops are grown into international conferences with many
hundreds of attendees, special issues are organized and published in a
leading scientific magazine, and the scope of the SciPy library
narrows, while the breadth of the ecosystem is grown and encouraged
through a new type of auxiliary package, the scikit.  The tooling,
development, and release processes become more professional, grown with
the patience affordable in open source development, leading with
best practices nowadays increasingly common in both industry and other
open source projects \cite{millman2014developing}.

@@

In the early workshops, recurrent topics indicate the state of development, with
emphasis being placed on the underlying array package, plotting,
parallel processing, acceleration / wrapping, and user interfaces.  By
2004, a significant shift occurred towards application of SciPy to
scientific problems.  The event also started to draw in more keynote
speakers from outside the community, such as Guido van Rossum (creator
of Python, 2006), Ivan Krstić (One Laptop per Child, 2007), Alex
Martelli (Google, 2008), and Peter Norvig (Google Research, 2009).
The SciPy conference went from being a small gathering of
core developers, to a multi-location event, with increased funding, a
published proceedings, and scholarships for attending students (some
of whom are authors on this paper).  In 2010, the conference grew even
further, with an additional event added in India, and the US
conference—now with parallel tracks—moving to Austin, Texas.
While the US SciPy conference kept growing (by ... it had multiple
tracks, student sponsorship, proceedings, and funding), satellite
events were being organized by volunteers elsewhere, such as EuroSciPy
(2008–) and SciPy India (2009–).
Special sessions and minisymposia dedicated to scientific Python began
appearing at many events.
For example, a three-part minisymposium organized for CSE 2009 was
featured in SIAM News\cite{siamcse09}.

When Paul edited the CiSE May/June 2007 special issue, Python already had a
strong enough presence in science and engineering so that editors solicited
the special issue, but it was still niche enough that the average reader
would need additional information in order to decide whether it would
``be useful in their own work.''
The follow-up CiSE March/April 2011 Python for Scientists and Engineers
special issue\cite{millman2011python}, focused more on the core parts
of the scientific Python ecosystem\cite{perez2011python} including
NumPy\cite{vanderwalt2011numpy}, Cython\cite{behnel2011cython},
and Mayavi\cite{ramachandran2011mayavi}.

Python has become so pervasive, there are now domain-specific special issues.
For example, Frontiers in Neuroinformatics has a collection of 25 articles---covering
topics including modeling and simulation, data collection, electrophysiology, visualization,
as well as stimulus generation and presentation---called Python in
neuroscience\footnote{\url{https://www.frontiersin.org/research-topics/8/python-in-neuroscience}}.

In 2012, Perry Greenfield, John Hunter, Jarrod Millman, Travis Oliphant,
and Fernando Pérez founded NumFOCUS\footnote{https://numfocus.org},
 a 501(c)3 public charity whose mission is to
``promote sustainable high-level programming languages, open code development,
and reproducible scientific research.''
While NumFOCUS is language agnostic, many of the early sponsored projects
came from the scientific Python stack.
Today it has over 25 sponsored projects including NumPy, SciPy, IPython, and
Matplotlib.
Among its other projects, NumFOCUS organizes a global network of 
community driven educational program called PyData.

@@

SciPy provides fundamental algorithms for scientific computing. The
breadth of its scope was derived from the Guide to Available Mathematical
Software classification system (GAMS\cite{boisvert1991guide}). In areas
that move relatively slowly, e.g. linear algebra, SciPy aims to provide
complete coverage. In other areas it aims to provide fundamental building
blocks while interacting well with other packages specialized in that area.
For example, SciPy provides what one expect to find in a
statistics textbook (probability distributions, hypothesis tests, frequency
statistics, correlation functions, and more), while Statsmodels provides
more advanced statistical estimators and inference methods, scikit-learn
covers machine learning, PyMC3, emcee and PyStan cover Bayesian statistics
and probabilistic modeling.

scikit-image

sympy\cite{meurer2017sympy}

NetworkX\cite{hagberg2008networkx}

@@

\subsection*{Current status (maturity, users)}

SciPy 1.0 was released ... see timeline ...

1.1 ...

1.2

Python 2 / 3?

\begin{figure}
\begin{verbatim}
    2001: first SciPy release
    2005: transition to NumPy
    2007: creation of scikits
    2008: spatial module and first Cython code added
    2010: move to a 6-monthly release cycle
    2011: development moves to GitHub
    2011: Python 3 support
    2012: add a sparse graph module and unified optimization interface
    2012: removal of maxentropy
    2013: continuous integration with TravisCI
    2015: adding Cython interface for BLAS/LAPACK and a benchmark suite
    2017: add a unified C API with LowLevelCallable; remove weave
    2017: SciPy 1.0 release
\end{verbatim}
\caption{Timeline}
\end{figure}

@@

The application programming interface (API) for SciPy consists of approximately
1500 functions and classes.  Our policy for evolving the API over time is that
new functionality can be added, while removing or changing existing
functionality can only be done if the benefits of that exceeds the (often
significant) costs to users, \textit{and} only after giving clear deprecation
warnings to those users for at least one year.  In general, we encourage
changes that improve clarity in the API of the library but strongly discourage
breaking backwards compatibility\footnote{
    Backwards compatibility means that new versions of SciPy will keep user
    code written for an older version working.}
given our position near the base of the scientific Python computing stack.

@@

Over the last three years our sparse matrix handling internals have been
rewritten and performance has been improved. Iterating over and slicing of CSC
and CSR matrices is now faster by up to 35\%, and the CSR to COO matrix format
conversion is now several times faster.  Importantly,
SuperLU\cite{superlu_ug99} was updated to version 5.2.1, enhancing the
low-level implementations leveraged by a subset of our \texttt{sparse}
offerings.

@@

The airspeed velocity (asv\cite{asvref}) library enables benchmarking Python packages
over their lifetimes, and the performance of the SciPy code base was monitored
with asv starting in February of 2015 (PR \#4501). In addition to ensuring that
unit tests are passing (see above), confirming that performance generally
remains constant or improves over the commit hash history of the project allows
us to objectively measure that our code base is improving, to empower
scientific applications.

Consider the asv benchmark results shown in Figure~\ref{fig:asvbench}, spanning
roughly nine years of project history. These demonstrate the gradual
performance improvements in a nearest-neighbor search through
\texttt{scipy.spatial.cKDTree.query()}, and can be run using the command:

\texttt{python run.py run -e -s 800 --bench "\textbackslash
btime\_query\textbackslash b" "02de46a546..b3ddb2c"}

@@

Any pull request can be compared against the \texttt{master} branch with the
command \texttt{asv continuous master new-feature}. This will provide a
benchmark against the master branch and the branch a new feature is implemented
in. More features are available in the documentation, including arguments to
select which benchmarks to run.

@@

>>>>

SciPy's arrival at this point is surprising, and somewhat anomalous.
When started, the library had little funding, and was written mainly
by graduate students---many of them without a computer science education, and often without the
blessing of their advisors.  To even imagine that a small group of
``rogue'' student programmers could upend the already well-established
ecosystem of research software---backed by millions in funding, and
many hundreds of highly qualified engineers\cite{mathworks-globe-97,
esri-revenue,bloom-wolfram}---was preposterous.

@@

In 1995, Jim Hugunin, a graduate student from MIT, wrote in the first
message on the Python Matrix Special Interest Group (Matrix-SIG)
mailing list\cite{Hugunin-first}:
\begin{quote}
There seems to be a fair amount of interest in the Python community
concerning the addition of numeric operations to Python.  My own desire is
to have as large a library of matrix based functions available as possible
(linear algebra, eigenfunctions, signal processing, statistics, etc.).  In
order to ensure that all of these libraries interoperate, there needs to
be agreement on a basic matrix object that can be used to represent arrays
of numbers.
\end{quote}
Over the next several months, conversations on that mailing
list by, among others, Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
Hinsen, and Guido van Rossum, led to the creation of an array object
that supported a high number of dimensions.  Jim Hugunin explained the utility
of Python for numerical computation\cite{Hugunin-whitepaper}:
\begin{quote}
I've used almost all of the available numerical languages at one time
or another over the past 8 years. One thing I've noticed is that over
time, the designers of these languages are steadily adding more of the
features that one would expect to find in a general-purpose
programming language.
\end{quote}
This is still a distinguishing feature of Python for science, and one of the
reasons why it has been so successful in the realm of data science: instead of
adding general features to a language designed for numerical and scientific
computing, here scientific features are added to a general purpose language
instead.  This broadens the scope of problems that can be addressed easily, the
sources of data that are readily accessible, and increases the size of the
community that develops general purpose tooling.

@@

Initial work on numerical computing in Python was driven by graduate
students, but soon larger research labs became increasingly engaged.
For example, Paul Dubois at Lawrence Livermore National Laboratory (LLNL) took over the
maintenance of Numeric and funded the writing of its
manual\cite{Numeric-manual}, and
%in the Program for Climate Model Diagnosis and Intercomparison at
%Lawrence Livermore National Laboratory built their Climate Data
%Analysis Tools (``an open-source set of Python modules and tools for
%accessing, manipulating, and plotting scientific data in general and
%climate data sets in
%particular''\footnote{https://mail.python.org/pipermail/numpy-discussion/2001-May/000645.html})
%on Numeric, the newly established Python array object.  Paul later
%became the maintainer for Numeric, and helped fund the writing of the
%first manual\footnote{https://conference.scipy.org/scipy2010/slides/travis_oliphant_keynote.pdf}
in 1998 the Space Telescope Science Institute (STScI), in charge of science
operations of the Hubble Space Telescope, decided to replace their
custom scripting language with Python\cite{STScI-slither}.

@@

% PhD Mayo Clinic, started faculty position at BYU in 2001
By the late 1990s, discussions appeared on Matrix-SIG,
expressing a desire for a complete scientific data analysis environment\cite{Travis-Keynote-2010}.
Travis Oliphant, a PhD student at the Mayo Clinic,
released a number of packages\cite{Travis-some-modules,Travis-enhance} 
that built on top of the Numeric array
package, and provided algorithms for signal processing, special
functions, sparse matrices, quadrature, optimization, Fast Fourier
Transforms, and more.  One of these packages, Multipack, was a set of
extension modules that wrapped Fortran and C libraries such as
ODEPACK, QUADPACK, and MINPACK, to solve and minimize nonlinear
equations, integrate differential equations, and fit splines (the
latter implemented by Pearu Peterson).  Robert Kern, then an
undergraduate student (and currently a SciPy core developer), provided
compilation instructions under Windows.
% 2001 PhD Tallinn Technical University
Around the same time, Pearu Peterson, a PhD student from Estonia,
released f2py\cite{Pearu-f2py-mail}, a command line tool for binding Python and Fortran
codes, and wrote modules for linear algebra and interpolation.
% Graduated 1999, Duke
Eric Jones, while a graduate student at Duke, wrote a number of
packages to support his dissertation, including a parallel job
scheduler and genetic optimizer.
% Graduated 1998, PhD from Brown, 97–99 postdoc at Harvard Medical
% School; at Harvard ever since
%
% https://mail.python.org/pipermail/scipy-dev/2001-November/000088.html
%
Gary Strangman, a postdoctoral fellow at Harvard Medical School,
published several descriptive and inferential statistical 
routines\cite{Strangman-modules}.

@@

With a rich programming environment and a numerical array object in
place, the time was ripe for the development of a full scientific
software stack.
In 2001, Eric Jones and Travis Vaught founded Enthought Scientific
Computing Solutions (now Enthought, Inc.) in Austin, Texas.  In an
effort to simplify the tool stack, they created the SciPy project,
centered around the SciPy library which would subsume all the
above-mentioned packages.
The new project quickly gained momentum, with a website and code
repository appearing in
February\cite{archived-scipyorg}, and
a mailing list announced in
June \cite{new-scipy-list}. By
August, a first release was announced\cite{first-scipy-rel}, an excerpt of which is shown in
Fig.~\ref{fig:announce-0.1}.
In September, the first documentation was
published\cite{first-scipy-docs}.
The first SciPy
workshop\cite{first-scipy-workshop}
was held in September 2002 at CalTech---a single track, two day event with 50
participants, many of them developers of SciPy and surrounding libraries.

@@

As SciPy, the algorithms library of the ecosystem, began attracting the attention
of large research organizations,
the next generation of graduate students and postdocs were already exploring
other aspects of the computational environment.
In 2000, Prabhu Ramachandran, a PhD student at the Indian Institute of
Technology, started work on a 3D visualization application, based on
Kitware's C++ Visualization Toolkit\cite{schroeder:2006:VTK}, called
Mayavi\cite{mayavi-intro}.
In 2001, Fernando Pérez, a graduate student at the University of
Colorado, Boulder, created the IPython interactive shell (that would
eventually blossom into Project Jupyter\cite{Kluyver:2016aa}), built into
IPP, an earlier attempt by Janko Hauser, and including code from
Nathan Gray's LazyPython.
% IPython hosted on SciPy:
% https://mail.python.org/pipermail/scipy-user/2003-March/001455.html
John Hunter, a postdoc
%John got his PhD in 2001, then a post-doc with John Milton (dept of
%Neurology, UChicago?) a couple of years; then research professor
%under Kurt Hecox, whereafter at Tradelink from 2006
at the University
of Chicago, liked the plotting functionality
available in MATLAB, but had problems accessing their laboratory's
license, which was governed by a hardware dongle.  In response, he
wrote a plotting library from scratch, and Matplotlib 0.1 was released
April 2003\cite{matplotlib-rel}.

@@

%In April 2006, Paul
%Dubois\footnote{https://mail.python.org/pipermail/numpy-discussion/2006-April/007487.html} wrote
%to the NumPy mailing list:
%\begin{quote}
%IEEE's magazine, Computing in Science and Engineering (CiSE), has asked me
%to put together a theme issue on the use of Python in Science and
%Engineering. I will write an overview to be accompanied by 3-5 articles [. . .]
%that show a diverse set of applications or
%tools, to give our readers a sense of whether or not Pyth%on might be useful
%in their own work.
%\end{quote}
In a May/June 2007 special issue of IEEE Computing in Science and
Engineering, Paul Dubois wrote\cite{dubois2007guest}:
\begin{quote}
LLNL now has many Python-based efforts built from scratch or wrapped around
legacy codes [. . .] hundreds of thousands
of lines of C++, Python, and Fortran 95, all working together just as we hoped,
doing compute-intensive calculations on massively parallel computers.
\end{quote}
While it is now more commonly accepted, Dubois took the time to explain to
the 2007 reader that ``interpreted doesn't mean slow or only interactive.''
He also shares his own interest in Python for ``computational steering''
where ``Python serves as the input language to a scientific application, and the actual
computations are performed both in Python itself and in compiled extensions.''
%https://www.computer.org/csdl/mags/cs/2007/03/c3007.html
Using Python for computational steering was one of the earliest uses of Python
in large scientific processing pipelines.
In addition to Dubois' overview there were also articles introducing
Python and SciPy \cite{oliphant2007python}, IPython \cite{perez2007ipython},
and  Matplotlib \cite{hunter2007matplotlib}.
There was also a diverse set of
research applications including
systems biology\cite{myers2007python},
astronomy\cite{greenfield2007reaching},
robotics\cite{krauss2007python},
nanophotonics\cite{bienstman2007python},
partial differential equations\cite{mardal2007using},
neuroimaging\cite{millman2007analysis},
geographic information systems\cite{shi2007python}, and
education\cite{backer2007computational, myers2007pythona}.

@@

By the middle to late 2000s, SciPy was starting to mature after a long phase of significant
growth and adoption.
The informal workshops grew into international conferences with many
hundreds of attendees. Special issues were organized and published in a
leading scientific journal\cite{dubois2007guest}, and the scope of the SciPy library
narrowed, while the breadth of the ecosystem grew
through a new type of auxiliary package, the scikit\cite{scikits-general}. 
The tooling, development, and release processes became more professional.
SciPy was expanded carefully, with the patience affordable in open source 
projects and via best practices common in industry \cite{millman2014developing}.

@@

In the early workshops, recurrent topics reflected the state of development, with
emphasis being placed on the underlying array package, plotting,
parallel processing, acceleration / wrapping, and user interfaces.  By
2004, a significant shift occurred towards application of SciPy to
scientific problems.  The event also started to draw in more keynote
speakers from outside the community, such as Guido van Rossum (creator
of Python, 2006), Ivan Krstić (One Laptop per Child, 2007), Alex
Martelli (Google, 2008), and Peter Norvig (Google Research, 2009).
The SciPy conference went from being a small gathering of
core developers, to a multi-location event, with increased funding, a
published proceedings, and scholarships for attending students (some
of whom are authors on this paper).  In 2010, the conference grew even
further, with an additional event added in India, and the US
conference—now with parallel tracks—moving to Austin, Texas.
While the US SciPy conference kept growing (by 2010 it had multiple
tracks, student sponsorship, proceedings, and funding), satellite
events were being organized by volunteers elsewhere, such as EuroSciPy
(2008–) and SciPy India (2009–).
Special sessions and minisymposia dedicated to scientific Python began
appearing at many events.
For example, a three-part minisymposium organized for CSE 2009 was
featured in SIAM News\cite{siamcse09}.

In 2007, Python had a strong enough presence in science and engineering
that the editors of IEEE Computing in Science and Engineering
(CiSE) solicited a special issue\cite{dubois2007guest}, edited by Paul
Dubois. However, Python was still sufficiently niche that the average reader
would need additional information in order to decide whether it would
``be useful in their own work.''
The follow-up CiSE March/April 2011 Python for Scientists and Engineers
special issue\cite{millman2011python}, focused more on the core parts
of the scientific Python ecosystem\cite{perez2011python} including
NumPy\cite{vanderwalt2011numpy}, Cython\cite{behnel2011cython},
and Mayavi\cite{ramachandran2011mayavi}.

Python has become so pervasive, there are now domain-specific special issues.
For example, Frontiers in Neuroinformatics has a collection of 25 articles---covering
topics including modeling and simulation, data collection, electrophysiology, visualization,
as well as stimulus generation and presentation---called Python in
neuroscience\cite{python-FIN}.

In 2012, Perry Greenfield, John Hunter, Jarrod Millman, Travis Oliphant,
and Fernando Pérez founded NumFOCUS\cite{numfocus},
 a 501(c)3 public charity whose mission is to
``promote sustainable high-level programming languages, open code development,
and reproducible scientific research.''
While NumFOCUS is language agnostic, many of the early sponsored projects
came from the scientific Python stack.
Today it has over 25 sponsored projects including NumPy, SciPy, IPython, and
Matplotlib.
Among its other projects, NumFOCUS organizes a global network of 
community driven educational programs, meetups, and conferences
called PyData.

@@

SciPy provides fundamental algorithms for scientific computing. The
breadth of its scope was derived from the Guide to Available Mathematical
Software classification system (GAMS\cite{boisvert1991guide}). In areas
that move relatively slowly, e.g. linear algebra, SciPy aims to provide
complete coverage. In other areas it aims to provide fundamental building
blocks while interacting well with other packages specialized in that area.
For example, SciPy provides what one expects to find in a
statistics textbook (probability distributions, hypothesis tests, frequency
statistics, correlation functions, and more), while 
Statsmodels\cite{statsmodels2010} provides
more advanced statistical estimators and inference methods;
scikit-learn\cite{pedregosa2011scikit} covers machine learning; and
PyMC3\cite{10.7717/peerj-cs.55}, emcee\cite{2013PASP-emcee} and 
PyStan\cite{pystan-ref} cover Bayesian statistics and probabilistic modeling.
scikit-image\cite{vanderwalt2014scikit} provides image processing
capabilities beyond \texttt{scipy.ndimage}, sympy\cite{meurer2017sympy}
provides a Python interface for symbolic computation, and
NetworkX\cite{hagberg2008networkx} is a Python library for probing
graphs and networks in a more specialized way than SciPy components
like \texttt{sparse.csgraph} or \texttt{spatial}.

@@

\subsection*{Current status}

At the time of writing, SciPy produces a new release
approximately every six months, and some of the major
feature highlights from the three years preceding
SciPy 1.0 are discussed in a technical improvements
section below. Some longer-term milestone events are
highlighted in Figure~\ref{fig:timeline}.

\begin{figure}[H]
\begin{verbatim}
    2001: first SciPy release
    2005: transition to NumPy
    2007: creation of scikits
    2008: spatial module and first Cython code added
    2010: move to a 6-monthly release cycle
    2011: development moves to GitHub
    2011: Python 3 support
    2012: add a sparse graph module and unified optimization interface
    2012: removal of maxentropy
    2013: continuous integration with TravisCI
    2015: adding Cython interface for BLAS/LAPACK and a benchmark suite
    2017: add a unified C API with LowLevelCallable; remove weave
    2017: SciPy 1.0 release
\end{verbatim}
\caption{Timeline}
\label{fig:timeline}
\end{figure}

@@

The application programming interface (API) for SciPy consists of approximately
1500 functions and classes.  Our policy for evolving the API over time is that
new functionality can be added, while removing or changing existing
functionality can only be done if the benefits of that exceeds the (often
significant) costs to users, \textit{and} only after giving clear deprecation
warnings to those users for at least one year. In general, we encourage
changes that improve clarity in the API of the library but strongly discourage
breaking backwards compatibility given our position near the base of the 
scientific Python computing stack.

@@

Over the last three years our sparse matrix handling internals have been
rewritten and performance has been improved. Iterating over and slicing of CSC
and CSR matrices is now faster by up to 35\% (Figure~\ref{fig:sparse-iter}),
 and the CSR to COO matrix format
conversion is now several times faster (Figure~\ref{fig:sparse-conv}).  Importantly,
SuperLU\cite{superlu_ug99} was updated to version 5.2.1, enhancing the
low-level implementations leveraged by a subset of our \texttt{sparse}
offerings.

@@

The airspeed velocity (asv\cite{asvref}) library enables benchmarking Python packages
over their lifetimes, and the performance of the SciPy code base was monitored
with asv starting in February of 2015 (PR \#4501). In addition to ensuring that
unit tests are passing (see Test suite section below), confirming 
that performance generally
remains constant or improves over the commit hash history of the project allows
us to objectively measure that our code base is improving, to empower
scientific applications.

Consider the asv benchmark results shown in Figure~\ref{fig:asvbench}, spanning
roughly nine years of project history. These demonstrate the gradual
performance improvements in a nearest-neighbor search through
\texttt{scipy.spatial.cKDTree.query()}, and can be run using:

\texttt{python run.py run -e -s 800 --bench "\textbackslash
btime\_query\textbackslash b" "02de46a546..b3ddb2c"},

where \texttt{run.py} is a light wrapper script around \texttt{asv}
that is packaged with SciPy.

@@

Any pull request can be compared against the \texttt{master} branch with the
command \texttt{asv continuous master new-feature}. This will provide a
benchmark against the master branch and the branch a new feature is implemented
in. More features are available in the \texttt{asv} documentation\cite{asvdocs}, including 
arguments to select which benchmarks to run.

@@

\section*{Supporting Information}

% separate Figure numbering for SI
\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{supporting_info/asv_bench/sparse/sparse_iteration_bench}
\caption{Airspeed velocity benchmarks for SciPy CSC and CSR matrix row iteration over roughly three years of commits prior to the release of version 1.0. Python version 2.7, Cython version 0.27.3, NumPy 1.8.2, and the indicated non-zero element density fractions were used. The values were obtained from our regularly-updated suite of performance results available online at \url{http://pv.github.io/scipy-bench/}.}
\label{fig:sparse-iter}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth]{supporting_info/asv_bench/sparse/sparse_csr_to_coo_bench}
\caption{Airspeed velocity benchmarks for SciPy CSR to COO matrix format conversion over roughly three years of commits prior to the release of version 1.0. Python version 2.7, Cython version 0.27.3, and NumPy 1.8.2 were used. The values were obtained from our regularly-updated suite of performance results available online at \url{http://pv.github.io/scipy-bench/}.}
\label{fig:sparse-conv}
\end{figure}
